---
title: "Croco_Colim_DESeq2"
author: "Nina Yang, PhD Candidate, Hutchins Lab, USC"
date: "9/28/2021"
updated: "3/10/2022"
output: html_document
---

## Background

This document outlines the steps taken to analyze the RNASeq data using DESeq2 

## Load libraries

**1. Install (if required) & upload packages into your current session**

```{r install-packages, eval = TRUE, message = FALSE, warning = FALSE}
if(!require(dplyr)){install.packages("dplyr")} 
if(!require(tidyverse)){install.packages("tidyverse")} 
if(!require(ggpubr)){install.packages("plyr")}
if(!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("DESeq2")
```

**2. Upload the package into your current session**
```{r library, eval = TRUE, message = FALSE, warning = FALSE}
library(DESeq2)
library(tidyverse) 
library(dplyr) 
library(plyr)
```

## RNASeq Analysis with DESeq2

Remove duplicate KOs

```{r ddply, eval = TRUE, message = FALSE, warning = FALSE}
# Gene counts were assigned to Kegg IDs and the duplicate KOs in the dataset will be merged.

df <- read.csv("keggassign_counts_fordeduping_stranded.csv") 
summary(df) # there are 5,924 entries
head(df)
dim(df)
ddDF <- ddply(df,"gene_id",numcolwise(sum))
dim(ddDF)
write.csv(ddDF,"CrocoColim_counts_deduped_forDESeq2.csv", row.names = FALSE)
```

Normalize data for subsequent analysis. After removing duplicates, the counts file will be trimmed to remove rows where the average of each treatment for a given gene is less than 5 (we keep genes where the average of at least one treatment is greater than 5). 

```{r trim, eval = TRUE, message = FALSE, warning = FALSE}
# trimmed counts is the data where genes with low counts have been removed
trim_df <- read_csv("CrocoColim_counts_deduped_forDESeq2.csv")
head(trim_df)
colnames(trim_df)

# calculate rowmeans
mean.CoD <- as.numeric(rowMeans(trim_df[, 2:4]))
mean.CoN <- as.numeric(rowMeans(trim_df[, 5:7]))
mean.FeD <- as.numeric(rowMeans(trim_df[, 8:10]))
mean.FeN <- as.numeric(rowMeans(trim_df[, 11:13]))
mean.PD <- as.numeric(rowMeans(trim_df[, 14:16]))
mean.PN <- as.numeric(rowMeans(trim_df[, 17:19]))
mean.ReD <- as.numeric(rowMeans(trim_df[, 20:22]))
mean.ReN <- as.numeric(rowMeans(trim_df[, 23:25]))

# create a dataframe only with rowmeans based on treatment
trim_df1 <- as.data.frame(cbind(mean.CoD, mean.CoN, mean.FeD, mean.FeN, mean.PD, mean.PN, mean.ReD, mean.ReN))
colnames(trim_df1)
summary(trim_df1)
trim_df2 <- cbind(trim_df$gene_id, trim_df1)
summary(trim_df2)
names(trim_df2)[1] <- "gene_id"
head(trim_df2)
summary(trim_df2)

# filter this rowMeans dataframe so that all rows where even one column is less than 5 is removed
trim_rows <- trim_df2 %>% dplyr::select(gene_id, starts_with("mean")) %>% filter_if(is.numeric, any_vars(. >= 5)) # should have 3770. Checked against a manual effort.

summary(trim_rows)

trimmed <- left_join(trim_rows, trim_df, by="gene_id")
head(trimmed)

final_trim <- trimmed %>% dplyr::select(gene_id, !(starts_with("mean")))
head(final_trim)
summary(final_trim)

write.csv(final_trim, file = "CrocoColim_trimmed_forDESeq2.csv", row.names = TRUE)

```

Setting up the matrix for DESeq2 and generating normalized datasets.

```{r normalize, eval = TRUE, message = FALSE, warning = FALSE}
cts <- read_csv("CrocoColim_trimmed_forDESeq2.csv")
colnames(cts);
cts <- cts[, 3:26]
colnames(cts);
colnames(cts)

# column data is the metadata for the dataset
## the data has Sample, Treatment, Diel, and Design (combination of Treatment and Diel)
col <- read.csv("colim_coldata_DESeq2.csv", header = TRUE)

# set up DESeq2 matrix for Design
dds <- DESeqDataSetFromMatrix(countData = cts, colData = col, design = ~ Design)

# get normalized counts: not for DESeq but important for downstream visualization (DESeq2 input is non-normalized counts data because it undergoes normalization as a part of the pipeline). We will be using normalized counts for visualizations of biomarker genes.
# https://hbctraining.github.io/DGE_workshop/lessons/02_DGE_count_normalization.html
# note that DESeq2 recommend vst transformed!!
ddsNorm <- estimateSizeFactors(dds)
sizeFactors(ddsNorm)
normalized_counts <- counts(ddsNorm, normalized=TRUE)
write.csv(as.data.frame(normalized_counts), file="CrocoColim_normcounts_forDGE.csv")

# plot a PCA to check the data 
vstcounts <- vst(dds, blind=FALSE)
write.csv((assay(vstcounts)), file="CrocoColim_vstcounts_forDGE.csv")
head(assay(vstcounts), 3)
plotPCA(vstcounts, intgroup=c("Treatment", "Diel"))
rld <- rlog(dds, blind=FALSE)
plotPCA(rld, intgroup=c("Treatment", "Design"))
write.csv((assay(rld)), file="CrocoColim_rldcounts_forDGE.csv")

# don't use normalized counts (norm counts are used for comparison and other calculations)

ddsObj <- DESeq(dds)
res <- results(ddsObj, alpha = 0.05) # filtering set to alpha 0.05
sum(res$padj < 0.05, na.rm=TRUE) # this pulls the number of genes with padj less than 0.05 (total) 
# 2,143

# Day v Night
## Day v Night Day means that positive is upregulated for Day relative to Night

#Replete
RepDN <- results(ddsObj, alpha = 0.05, contrast = c("Design", "RepD", "RepN")) 
summary(RepDN)

# write the file as a csv
write.csv(as.data.frame(RepDN), 
          file="RepDN_DESeqresults.csv")

#Colim
CoDN <- results(ddsObj, alpha = 0.05, contrast = c("Design", "CoD", "CoN")) 
summary(CoDN)

# write the file as a csv
write.csv(as.data.frame(CoDN), 
          file="CoDN_DESeqresults.csv")

#Felim
FeDN <- results(ddsObj, alpha = 0.05, contrast = c("Design", "FeD", "FeN")) 
summary(FeDN)

# write the file as a csv
write.csv(as.data.frame(FeDN), 
          file="FeDN_DESeqresults.csv")

#Plim
PDN <- results(ddsObj, alpha = 0.05, contrast = c("Design", "PhD", "PhN")) 
summary(PDN)

# write the file as a csv
write.csv(as.data.frame(PDN), 
          file="PDN_DESeqresults.csv")

# Colim vs Replete Day
## Colim_Day v Replete_Day means that positive is upregulated for Co relative to Rep
resCoRepDay <- results(ddsObj, alpha = 0.05, contrast = c("Design", "CoD", "RepD")) 
summary(resCoRepDay)

# write the file as a csv
write.csv(as.data.frame(resCoRepDay), 
          file="CoRepDay_DESeqresults.csv")

# Felim vs Replete Day
resFeRepDay <- results(ddsObj, alpha = 0.05, contrast = c("Design", "FeD", "RepD"))
## extracts only Felim Day vs Replete Day results
## Felim_Day v Replete_Day means that positive is upregulated for Fe relative to Rep
summary(resFeRepDay)

# write the file as a csv
write.csv(as.data.frame(resFeRepDay), 
          file="FeRepDay_DESeqresults.csv")

# Plim vs Replete Day
resPRepDay <- results(ddsObj, alpha = 0.05, contrast = c("Design", "PhD", "RepD")) 
## extracts only Plim Day vs Replete Day results
## Plim_Day v Replete_Day means that positive is upregulated for P relative to Rep
summary(resPRepDay)

# write the file as a csv
write.csv(as.data.frame(resPRepDay), 
          file="PRepDay_DESeqresults.csv")


####### DESeq2 Night v Replete #######

# Colim vs Replete Night
resCoRepNite <- results(ddsObj, alpha = 0.05, contrast = c("Design", "CoN", "RepN")) 
## extracts only Colim Night vs Replete Night results
## Colim_Night v Replete_Night means that positive is upregulated for Co relative to Rep
summary(resCoRepNite)

# write the file as a csv
write.csv(as.data.frame(resCoRepNite), 
          file="CoRepNite_DESeqresults.csv")

# Felim vs Replete Night
resFeRepNite <- results(ddsObj, alpha = 0.05, contrast = c("Design", "FeN", "RepN")) 
## extracts only Felim Night vs Replete Night results
## Felim_Night v Replete_Night means that positive is upregulated for Fe relative to Rep
summary(resFeRepNite)

# write the file as a csv
write.csv(as.data.frame(resFeRepNite), 
          file="FeRepNite_DESeqresults.csv")

# Plim vs Replete Night
resPRepNite <- results(ddsObj, alpha = 0.05, contrast = c("Design", "PhN", "RepN")) 
## extracts only Plim Night vs Replete Night results
## Plim_Night v Replete_Night means that positive is upregulated for P relative to Rep
summary(resPRepNite)

# write the file as a csv
write.csv(as.data.frame(resPRepNite), 
          file="PRepNite_DESeqresults.csv")


# Felim vs Colim Day
resFeCoDay <- results(ddsObj, alpha = 0.05, contrast = c("Design", "FeD", "CoD")) 
## extracts only Plim Night vs Replete Night results
## Plim_Night v Replete_Night means that positive is upregulated for P relative to Rep
summary(resFeCoDay)

# write the file as a csv
write.csv(as.data.frame(resFeCoDay), 
          file="FeCoDay_DESeqresults.csv")

# Felim vs Colim Night
resFeCoNite <- results(ddsObj, alpha = 0.05, contrast = c("Design", "FeN", "CoN")) 
summary(resFeCoNite)

# write the file as a csv
write.csv(as.data.frame(resFeCoNite), 
          file="FeCoNite_DESeqresults.csv")

# Plim vs Colim Day

resPCoDay <- results(ddsObj, alpha = 0.05, contrast = c("Design", "PhD", "CoD")) 
summary(resPCoDay)

# write the file as a csv
write.csv(as.data.frame(resPCoDay), 
          file="PCoDay_DESeqresults.csv")

# Plim vs Colim Night

resPCoNite <- results(ddsObj, alpha = 0.05, contrast = c("Design", "PhN", "CoN")) 
summary(resPCoNite)

# write the file as a csv
write.csv(as.data.frame(resPCoNite), 
          file="PCoNight_DESeqresults.csv")
```
